\begin{thebibliography}{99}

\bibitem[Collins 2004]{collins04} Collins-Sussman, B.: Version Control with Subversion.
\textit{Online http://svnbook.red-bean.com/svnbook/book.html}, May 2004.

\bibitem[Dourish 1996]{dourish96} Dourish, P.: Consistency guarantees: Exploiting application semantics for consistency management in a collaboration toolkit.
\textit{Proceedings of the 1996 ACM Conference on Computer-Supported Cooperative Work}, Boston, USA, 1996, pp. 268-277.

\bibitem[Ignat 2002]{ignat02} Ignat, C., Norrie, M.: Tree-based model algorithm for maintaining consistency in real-time collaborative editing systems.
\textit{The 4th International Workshop on Collaborative Editing, Computer Supported Collaborative Work}, New Orleans, USA, Nov 2002.

\bibitem[Ignat 2003]{ignat03} Ignat, C., Norrie, M.: Customizable collaborative editor relying on treeOPT algorithm.
\textit{Proceedings of the 8th European Conference on Computer Supported Collaborative Work}, Helsinki, Finland, Sep 2003, pp. 315-334.

\bibitem[Ignat 2004a]{ignat04a} Ignat, C., Norrie, M.: Extending real-time editing systems with asynchronous communication.
\textit{Proceedings of the 8th International Conference on Computer Supported Collaborative Work in Design}, Xiamen, P. R. China, May 2004.

\bibitem[Ignat 2004b]{ignat04b} Ignat, C., Norrie, M.: CoDoc: Multi-mode collaboration over documents.
\textit{Proceedings of the 16th International Conference on Advanced Information Systems Engineering}, Riga, Latvia, Jun 2004.

\bibitem[Imine et al. 2003]{imine03} Imine, A., Molli, P., Oster, G., Rusinowitch, M.: Proving correctness of transformation functions in real-time groupware.
\textit{Proceedings of the 8th European Conference on Computer-Supported Cooperative Work}, Helsinki, Finland, Sep 2003.

\bibitem[Ionescu and Marsic 2000]{ion00} Ionescu, M., Marsic, I.: An arbitration scheme for concurrency control in distributed groupware.
\textit{Proceedings of the 2nd International Workshop on Collaborative Editing Systems, An ACM CSCW'2000 Workshop}, Philadelphia, USA, Dec 2000.

\bibitem[Molli et al. 2002]{molli02} Molli, P., Skaf-Molli, H., Oster, G., Jourdain, S.:  Sams: Synchronous, asynchronous, multi-synchronous environments.
\textit{The 7th International Conference on Computer Supported Collaborative Work in Design}, Rio de Janeiro, Brazil, Sep 2002.

\bibitem[Molli et al. 2003]{molli03} Molli, P., Oster, G., Skaf-Molli, H., Imine, A.: Using the transformational approach to build a safe and generic data synchronizer.
\textit{Proceedings of the 2003 International ACM SIGGROUP conference on Supporting group work}, Sanibel Island, Florida, USA, pp. 212-220.

\bibitem[Munson 1994]{munson94} Munson, J. P., Dewan, P.: A flexible object merging framework.
\textit{Proceedings of the 1994 ACM Conference on Computer-Supported Cooperative Work}, Chapel Hill, USA, 1994, pp. 231-242.

\bibitem[Nedevschi 02]{ned02} Nedevschi, S.: Concurrency control in real-time collaborative editing systems
\textit{Diploma thesis}, Z\"{u}rich, Switzerland, June 2002.

\bibitem[Shen and Sun 2002]{shen02} Shen, H., Sun, C.: Flexible merging for asynchronous collaborative systems.
\textit{Proceedings of the 10th International Conference on Cooperative Information Systems}, Springer Verlag, USA, Nov 2002, pp. 304-321.

\bibitem[Sun et al. 1996]{sun96} Sun, C., Yang, Y., Zhang, Y., Chen, D.: A consistency model and supporting schemes for real-time cooperative editing systems.
\textit{Proceedings of the 19th Australian Computer Science Conference}, Melbourne, Jan 1996, pp. 582-591.

\bibitem[Sun et al. 1998]{sun98a} Sun, C., Jia, X., Zhang, Y., Yang, Y., Chen, D.: Achieving convergence, causality-preservation, and intention-preservation in real-time cooperative editing systems.
\textit{ACM Transactions on Computer-Human Interaction}, vol. 5, no. 1, Mar 1998, pp. 63-108.

\bibitem[Sun and Ellis 1998]{sun98b} Sun C., Ellis, C.: Operational transformation in real-time group editors: issues, algorithms, and achievements.
\textit{Proceedings of ACM Conference on Computer Supported Cooperative Work}, Seattle, USA, Nov 1998, pp. 59-68.

\bibitem[Bitkeeper Home Page]{bitk} Bitkeeper home page.
\textit{Online http://www.bitkeeper.com/}, May 2004.

\bibitem[VSS Home Page]{vss} Microsoft Visual SourceSafe home page.
\textit{Online http://msdn.microsoft.com/vstudio/previous/ssafe/}, May 2004.

\bibitem[CVS]{cvs} Concurrent Versions System.
\textit{Online http://en.wikipedia.org/wiki/Concurrent\_versions\_system}, May 2004.

\bibitem[Subversion Home Page]{subv} Subversion home page.
\textit{Online http://subversion.tigris.org/}, May 2004.

\end{thebibliography}